---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The zipped csv file is first loaded with the columns defined as:

* steps - type 'numeric'
* date - type 'Date'
* interval - type 'numeric'

```{r}
rm(list=ls())
setwd('H:/Courses/DataScience/ReproducibleResearch/Assignment1/RepData_PeerAssessment1')
fileName = 'activity.zip'
data <- read.csv(unz(fileName, "activity.csv"),colClasses = c('numeric','Date','numeric'))
```

The data loaded has the following format

```{r,results='asis',warning=FALSE}
library(pander)
pander(data[seq(1,dim(data)[1],length.out = 6),],style='rmarkdown')
```

As can be seen there are *NA* values present. The effects of excluding and imputing the missing values will be explored in the course of this document.

## What is the average total number of steps taken per day?

For this part of the assignment, **the missing values in the dataset will be ignored**. A histogram of the total number of steps taken each day will be shown. The average and median of the total number of steps taken per day will be reported.

```{r,message=FALSE,fig.align='center',warning=FALSE}
require(dplyr)
require(ggplot2)

# Steps: filter by excluding NA values, then group by date and sum each date by summarising
stepsByDate=filter(data,!is.na(steps)) %>% select(-interval) %>% group_by(date)
totalStepsByDate=summarise(stepsByDate,totalSteps=sum(steps))
summaryStats=summarise(totalStepsByDate,MEAN=mean(totalSteps),MEDIAN=median(totalSteps))

# Set the bin width of the histogram so we avoid 'gaps'
binw=diff(range(totalStepsByDate$totalSteps))/7

# And plot
g <- ggplot(data=totalStepsByDate,aes(x=totalSteps,fill=..count..))
g+scale_fill_gradient(guide='none')+
        theme_bw()+
        xlab('Total steps per day')+
        ylab('Frequency')+
        ggtitle('Histogram of total number of steps each day')+
        geom_histogram(binwidth=binw, colour='black')
```

Some summary statistics:

* The **average** total number of steps every day is **`r sprintf('%1.0f',summaryStats$MEAN)`**.
* The **median** total number of steps every day is **`r sprintf('%1.0f',summaryStats$MEDIAN)`**.

> Note that the average and median in this case is very close to each other. This result, however,
> depends on no assumptions around missing data - *NA* points will simply not be considered. Later on 
> the impact on the distribution, when missing values are imputed, will be considered.

## What is the average daily activity pattern?

In this section the time series plot of 

* the 5-minute interval (x-axis) and the 
* average number of steps taken, averaged across all days (y-axis) 

will be made. The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps will be reported.

```{r, fig.align='center'}

# remove NAs, group by each interval and get the mean
stepsByInterval <- filter(data,!is.na(steps)) %>% group_by(interval) %>%
                        summarise(averageSteps=mean(steps))

# determine the max point and store that row
maxDets <- filter(stepsByInterval,averageSteps==max(averageSteps))

# plot
g <- ggplot(data=stepsByInterval,aes(x=interval,y=averageSteps))
g+theme_bw()+
        ylab('Average number of steps over all days')+
        ggtitle('Average steps over all days')+
        geom_area(fill='#6495ED')+
        geom_line(colour='black')+
        geom_text(data=maxDets,aes(x=interval,y=averageSteps,
                  label=paste('MAX','=',round(averageSteps,0),'steps','at the',
                  interval,'minute interval')),color='darkred',vjust=-0.01,hjust=-0.01)
```

The interval containing the maximum averaged steps is `r sprintf('%1.0f',maxDets$interval)`.

## Imputing missing values

The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r}
totMissing=select(data,steps) %>% filter(is.na(steps)) %>% summarise(n())
```

The total number of missing values in the dataset is **`r sprintf('%1.0f',totMissing)`**.

A possible strategy to impute the missing data is to calculate the median of the number of steps by 

* weekday (to allow for day of the week effects) and
* interval (to allow for time of day effects)

simultaneously.

```{r}
# add the weekday and calculate the median by weekday and interval
addByDay=mutate(data,DAYOFWEEK=weekdays(date))
medianByDay <- filter(addByDay,!is.na(steps))%>% 
        group_by(DAYOFWEEK,interval) %>%
        summarise(MEDIAN=median(steps))

# Now replace the missing values
newSteps <- sapply(1:dim(addByDay)[1],function(i){
        x=addByDay[i,]
        if(is.na(x['steps']))
        {
                replRow=filter(medianByDay,DAYOFWEEK==x['DAYOFWEEK'] & interval==x['interval'])
                return(replRow['MEDIAN'])
        }else
        {
                return(x['steps'])
        }
})

newSteps<-unlist(newSteps)
addByDay <- mutate(addByDay,steps=newSteps)
```

The missing datapoints are now populated. For example, consider the same rows as in the original dataset:

```{r,results='asis'}
pander(addByDay[seq(1,dim(addByDay)[1],length.out = 6),],style='rmarkdown')
```

Next a histogram of the total number of steps taken each day will be made. The average 
and median total number of steps taken per day will be calculated. 

```{r,fig.align='center'}
# histogram and stats
stepsByDate=filter(addByDay,!is.na(steps)) %>% select(-interval) %>% group_by(date)
totalStepsByDate=summarise(stepsByDate,totalSteps=sum(steps))

#calculate the mean and median with imputed values
summaryStatsImputed=summarise(totalStepsByDate,MEAN=mean(totalSteps),MEDIAN=median(totalSteps))

# set bins to minimise 'gaps'
binw=diff(range(totalStepsByDate$totalSteps))/7

# and plot
g <- ggplot(data=totalStepsByDate,aes(x=totalSteps,fill=..count..))
g+scale_fill_gradient(guide='none')+
        theme_bw()+
        xlab('Total steps per day')+
        ylab('Frequency')+
        ggtitle('Histogram of total number of steps each day (including Imputed)')+
        geom_histogram(binwidth=binw, colour='black')
```

Summary statistics with imputed values:

* The **average** total number of steps every day is **`r sprintf('%1.0f',summaryStatsImputed$MEAN)` (previously `r sprintf('%1.0f',summaryStats$MEAN)`)**.
* The **median** total number of steps every day is **`r sprintf('%1.0f',summaryStatsImputed$MEDIAN)` (previously `r sprintf('%1.0f',summaryStats$MEDIAN)`)**.

> As can be seen, the inclusion of imputed data increases the left tail of the distribution.
> This has the effect of moving the average (which will be more sensitive to the tail change)
> more left relative to the median.

## Are there differences in activity patterns between weekdays and weekends?

First, the day of the week will be added. The business days / weekends will be separated by adding
a column consisting of two levels - "weekday" and "weekend".

```{r}
# Weekday already added. Add business day/weekend factor
addByDay<-mutate(addByDay,WKDAY=factor(ifelse((DAYOFWEEK!='Saturday')&(DAYOFWEEK!='Sunday'),
                                      'weekday','weekend')))
```

For example, the dataset now looks as follows:

```{r,results='asis'}
pander(addByDay[seq(1,dim(addByDay)[1],length.out = 6),],style='rmarkdown')
```

Next a panel plot containing the time series plot of 

* the 5-minute interval and the average number of steps taken, 
* averaged across all weekday days or weekend days (y-axis) 

will be made.

```{r,fig.align='center'}

# Average by each factor and interval
avgweekdayAndweekend <- group_by(addByDay,interval,WKDAY) %>%summarise(AVGSTEPS=mean(steps))

# And plot
g <- ggplot(data=avgweekdayAndweekend,aes(x=interval,y=AVGSTEPS))
g+theme_bw()+
        theme(strip.text.y = element_text(size = 15, colour = "yellow",angle = 90),
              strip.background=element_rect(fill='black'))+
        ylab('Average number of steps over all days')+
        ggtitle('Average steps over all days')+
        geom_area(fill='#6495ED')+
        geom_line(colour='black')+facet_grid(WKDAY ~.)
```